Balsa: A Compact Line Notation Based on SMILES

### 2023-02-05 manuscript.tex
States EBNF variant and references it. Closes #59
Removes unneeded grammar quantifier.
Adds "no trailing branch" restriction."

Balsa's formal grammar is presented as a series of \textit{production rules} (\enquote{productions}). A production defines a transformation allowed under the grammar. These transformations collectively define the set of valid Balsa strings. Productions can be used in the forward direction, when writing a string, or in the reverse direction, when reading a string.

@@

\begin{lstlisting}
branch ::= "(" ( dot | bond )? sequence ")"
           ( union | branch)+
\end{lstlisting}

@@

ProtoSMILES defines a recursive grammar for branching which, perhaps without the knowledge of the author, allows sequentially nested parentheses~\cite[p.~86]{weininger:2003}. Balsa disallows such constructs (e.g., \enquote{*((*))*} and \enquote{*(((*)))*}).

@@

\begin{table}
\begin{minipage}{\textwidth}
\renewcommand*\footnoterule{}
    \caption{Some differences between ProtoSMILES and Balsa.}
    \centering
    \begin{tabular}{p{0.50\linewidth} p{0.20\linewidth} p{0.15\linewidth}}
        \hline
        Feature & ProtoSMILES\footnote{Values of \enquote{maybe} indicate underspecified and/or internally inconsistent features.} & Balsa \\
        \hline
        element symbol Ha & yes & no \\
        multiple branching (e.g., \enquote{*((*))*}) & yes & no \\
        reactions using greater than symbol (\ttt{>}) & yes & no \\
        atomic \enquote{map} attribute & yes & no \\
        extended stereodescriptors (e.g., \enquote{@OH1}) & yes & no \\
        use of stereodescriptors on odd cumulene centers & yes & no \\
        virtual hydrogen count on hydrogen (e.g., \enquote{[HH]}) & maybe & yes \\
        gaps are bonds of \enquote{formal order zero} & maybe & no \\
        unbounded numerical atomic properties & yes & no \\
        atoms associated with any element are selectable & yes & no \\
        non-elided bonds may be promoted during selection & maybe & no \\
        isolated partial parity bond & yes & no \\
        repeated charge notation (e.g., \enquote{++} and \enquote{-{}-}) & yes & no \\
        aromatic bond (\enquote{:}) & yes & no \\
        \hline
    \end{tabular}
    \label{table:feature-comparison}
\end{minipage}
\end{table}

>>>>

Balsa's formal grammar is presented as a series of \textit{production rules} (\enquote{productions}). A production defines a transformation allowed under the grammar. These transformations collectively define the set of valid Balsa strings. Productions can be used in the forward direction, when writing a string, or in the reverse direction, when reading a string. Productions are expressed using the variation of Extended Backus-Naur Form defined by the World Wide Web Consortium~\cite{xmlspec}.

@@

\begin{lstlisting}
branch ::= "(" ( dot | bond )? sequence ")"
           ( union | branch)
\end{lstlisting}

@@

ProtoSMILES defines a recursive grammar for branching which, perhaps without the knowledge of the author, allows sequentially nested parentheses~\cite[p.~86]{weininger:2003}. Balsa disallows such constructs (e.g., \enquote{*((*))*} and \enquote{*(((*)))*}). Eliminating this kind of purposeless syntax simplifies software implementation. For the same reason, Balsa disallows trailing branches (e.g., \enquote{*(*)}).

@@

\begin{table}
\begin{minipage}{\textwidth}
\renewcommand*\footnoterule{}
    \caption{Some differences between ProtoSMILES and Balsa.}
    \centering
    \begin{tabular}{p{0.50\linewidth} p{0.20\linewidth} p{0.15\linewidth}}
        \hline
        Feature & ProtoSMILES\footnote{Values of \enquote{maybe} indicate underspecified and/or internally inconsistent features.} & Balsa \\
        \hline
        element symbol Ha & yes & no \\
        multiple branching (e.g., \enquote{*((*))*}) & yes & no \\
        trailing branch (e.g., \enquote{*(*)}) & yes & no \\
        reactions using greater than symbol (\ttt{>}) & yes & no \\
        atomic \enquote{map} attribute & yes & no \\
        extended stereodescriptors (e.g., \enquote{@OH1}) & yes & no \\
        use of stereodescriptors on odd cumulene centers & yes & no \\
        virtual hydrogen count on hydrogen (e.g., \enquote{[HH]}) & maybe & yes \\
        gaps are bonds of \enquote{formal order zero} & maybe & no \\
        unbounded numerical atomic properties & yes & no \\
        atoms associated with any element are selectable & yes & no \\
        non-elided bonds may be promoted during selection & maybe & no \\
        isolated partial parity bond & yes & no \\
        repeated charge notation (e.g., \enquote{++} and \enquote{-{}-}) & yes & no \\
        aromatic bond (\enquote{:}) & yes & no \\
        \hline
    \end{tabular}
    \label{table:feature-comparison}
\end{minipage}
\end{table}
